---
title: "Cancer of the Oral Cavity"
author: "René D Herrera"
date: "6/4/2020"
output:
  word_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
# set up ----
# load packages
library(here)
library(tidyverse)
library(knitr)
library(ggthemes)

# arizona, all races ----
# age adjusted incidence for cancer of the oral cavity
# read dataset to environment
# years 1997-2017
# cancer site: oral cavity
# all races
# all AZ counties
# both male and female sex
oral_cavity <- read_csv("../../../data/raw/az_cancer_registry_incidence_oral_cavity_20200604.csv",
  na = c("", "NA", "**", "*"),
  skip = 1,
  col_names = c(
    "Year",
    "Sex",
    "Case_Count",
    "Population",
    "Age_Adj_Rate",
    "95CI_min",
    "95CI_max"
  ),
  col_types = cols(
    Year = col_factor(),
    Sex = col_factor(),
    Case_Count = col_number(),
    Population = col_number(),
    Age_Adj_Rate = col_number(),
    "95CI_min" = col_number(),
    "95CI_max" = col_number()
  ),
)

glimpse(oral_cavity)

# calculate crude rate and add calculated variable
oral_cavity <- oral_cavity %>%
  mutate(crude_rate = round((Case_Count / Population) * 100000, digits = 2))

# generate plot visualization
AZ_oral_cavity_incidence_plot <- oral_cavity %>%
  filter(Year != "All") %>%
  ggplot(mapping = aes(y = Age_Adj_Rate, x = Year, color = Sex)) +
  geom_point() +
  scale_y_continuous(limits = c(0, 20)) +
  scale_x_discrete(labels = NULL) +
  labs(
    title = "Age-Adjusted Cancer Incidence Rates: Oral Cavity",
    subtitle = "for Arizona, years 1997-2017",
    y = "Age Adjusted Rate per 100,000",
    x = "Year 1997-2017",
    caption = "Source: Arizona Cancer Registry, 1997-2017"
  ) +
  theme_solarized()

# prepare data table with variables Year, Male, Female, Total
AZ_oral_cavity_incidence_table <- oral_cavity %>%
  filter(Year != "All") %>%
  select(Year, Sex, Age_Adj_Rate) %>%
  spread(Sex, Age_Adj_Rate) %>%
  select(Year, Male, Female, All) %>%
  kable()

# arizona, white non-hispanic ----
# age adjusted incidence for cancer of the oral cavity
# read dataset to environment
# years 1997-2017
# cancer site: oral cavity
# white, non-hispanic
# all AZ counties
# both male and female sex
oral_cavity_white <- read_csv("../../../data/raw/az_cancer_registry_incidence_oral_cavity_white_by_year_20200604.csv",
  na = c("", "NA", "**", "*"),
  skip = 1,
  col_names = c(
    "Year",
    "Sex",
    "Case_Count",
    "Population",
    "Age_Adj_Rate",
    "95CI_min",
    "95CI_max"
  ),
  col_types = cols(
    Year = col_factor(),
    Sex = col_factor(),
    Case_Count = col_number(),
    Population = col_number(),
    Age_Adj_Rate = col_number(),
    "95CI_min" = col_number(),
    "95CI_max" = col_number()
  ),
)

glimpse(oral_cavity_white)

# calculate crude rate and add calculated variable
oral_cavity_white <- oral_cavity_white %>%
  mutate(crude_rate = round((Case_Count / Population) * 100000, digits = 2))

# generate plot visualization
AZ_oral_cavity_incidence_white_plot <- oral_cavity_white %>%
  filter(Year != "All") %>%
  ggplot(mapping = aes(y = Age_Adj_Rate, x = Year, color = Sex)) +
  geom_point() +
  scale_y_continuous(limits = c(0, 20)) +
  scale_x_discrete(labels = NULL) +
  labs(
    title = "Age-Adjusted Cancer Incidence Rates: Oral Cavity",
    subtitle = "for Arizona, years 1997-2017, White, Non-Hispanic",
    y = "Age Adjusted Rate per 100,000",
    x = "Year 1997-2017",
    caption = "Source: Arizona Cancer Registry, 1997-2017"
  ) +
  theme_solarized()

# prepare data table with variables Year, Male, Female, Total
AZ_oral_cavity_incidence_white_table <- oral_cavity_white %>%
  filter(Year != "All") %>%
  select(Year, Sex, Age_Adj_Rate) %>%
  spread(Sex, Age_Adj_Rate) %>%
  select(Year, Male, Female, All) %>%
  kable()

# arizona, white hispanic ----
# age adjusted incidence for cancer of the oral cavity
# read dataset to environment
# years 1997-2017
# cancer site: oral cavity
# white, Hispanic
# all AZ counties
# both male and female sex
oral_cavity_hispanic <- read_csv("../../../data/raw/az_cancer_registry_incidence_oral_cavity_hispanic_by_year_20200604.csv",
  na = c("", "NA", "**", "*"),
  skip = 1,
  col_names = c(
    "Year",
    "Sex",
    "Case_Count",
    "Population",
    "Age_Adj_Rate",
    "95CI_min",
    "95CI_max"
  ),
  col_types = cols(
    Year = col_factor(),
    Sex = col_factor(),
    Case_Count = col_number(),
    Population = col_number(),
    Age_Adj_Rate = col_number(),
    "95CI_min" = col_number(),
    "95CI_max" = col_number()
  ),
)

glimpse(oral_cavity_hispanic)

# calculate crude rate and add calculated variable
oral_cavity_hispanic <- oral_cavity_hispanic %>%
  mutate(crude_rate = round((Case_Count / Population) * 100000, digits = 2))

# generate plot visualization
AZ_oral_cavity_incidence_hispanic_plot <- oral_cavity_hispanic %>%
  filter(Year != "All") %>%
  ggplot(mapping = aes(y = Age_Adj_Rate, x = Year, color = Sex)) +
  geom_point() +
  scale_y_continuous(limits = c(0, 20)) +
  scale_x_discrete(labels = NULL) +
  labs(
    title = "Age-Adjusted Cancer Incidence Rates: Oral Cavity",
    subtitle = "for Arizona, years 1997-2017, White, Hispanic",
    y = "Age Adjusted Rate per 100,000",
    x = "Year 1997-2017",
    caption = "Source: Arizona Cancer Registry, 1997-2017"
  ) +
  theme_solarized()

# prepare data table with variables Year, Male, Female, Total
AZ_oral_cavity_incidence_hispanic_table <- oral_cavity_hispanic %>%
  filter(Year != "All") %>%
  select(Year, Sex, Age_Adj_Rate) %>%
  spread(Sex, Age_Adj_Rate) %>%
  select(Year, Male, Female, All) %>%
  kable()

# catchment, all races ----
# age adjusted incidence for cancer of the oral cavity
# read dataset to environment
# years 1997-2017
# cancer site: oral cavity
# all races
# Cochise, Pima, Pinal, Santa Cruz, and Yuma counties only
# both male and female sex
oral_cavity_catch <- read_csv("../../../data/raw/az_cancer_registry_incidence_oral_cavity_catchment_20200604.csv",
  na = c("", "NA", "**", "*"),
  skip = 1,
  col_names = c(
    "Year",
    "Sex",
    "Case_Count",
    "Population",
    "Age_Adj_Rate",
    "95CI_min",
    "95CI_max"
  ),
  col_types = cols(
    Year = col_factor(),
    Sex = col_factor(),
    Case_Count = col_number(),
    Population = col_number(),
    Age_Adj_Rate = col_number(),
    "95CI_min" = col_number(),
    "95CI_max" = col_number()
  ),
)

glimpse(oral_cavity_catch)

# calculate crude rate and add calculated variable
oral_cavity_catch <- oral_cavity_catch %>%
  mutate(crude_rate = round((Case_Count / Population) * 100000, digits = 2))

# generate plot visualization
AZ_catch_oral_cavity_plot <- oral_cavity_catch %>%
  filter(Year != "All") %>%
  ggplot(mapping = aes(y = Age_Adj_Rate, x = Year, color = Sex)) +
  geom_point() +
  scale_y_continuous(limits = c(0, 20)) +
  scale_x_discrete(labels = NULL) +
  labs(
    title = "Age-Adjusted Cancer Incidence Rates: Oral Cavity",
    subtitle = "for Catchment, years 1997-2017",
    y = "Age Adjusted Rate per 100,000",
    x = "Year 1997-2017",
    caption = "Source: Arizona Cancer Registry, 1997-2017"
  ) +
  theme_solarized()

# prepare data table with variables Year, Male, Female, Total
AZ_catch_oral_cavity_table <- oral_cavity_catch %>%
  filter(Year != "All") %>%
  select(Year, Sex, Age_Adj_Rate) %>%
  spread(Sex, Age_Adj_Rate) %>%
  select(Year, Male, Female, All) %>%
  kable()

# catchment, white non-hispanic ----
# age adjusted incidence for cancer of the oral cavity
# read dataset to environment
# years 1997-2017
# cancer site: oral cavity
# white, non-hispanic
# Cochise, Pima, Pinal, Santa Cruz, and Yuma counties only
# both male and female sex
oral_cavity_catch_white <- read_csv("../../../data/raw/az_cancer_registry_incidence_oral_cavity_catchment_white_by_year_20200604.csv",
  na = c("", "NA", "**", "*"),
  skip = 1,
  col_names = c(
    "Year",
    "Sex",
    "Case_Count",
    "Population",
    "Age_Adj_Rate",
    "95CI_min",
    "95CI_max"
  ),
  col_types = cols(
    Year = col_factor(),
    Sex = col_factor(),
    Case_Count = col_number(),
    Population = col_number(),
    Age_Adj_Rate = col_number(),
    "95CI_min" = col_number(),
    "95CI_max" = col_number()
  ),
)

glimpse(oral_cavity_catch_white)

# calculate crude rate and add calculated variable
oral_cavity_catch_white <- oral_cavity_catch_white %>%
  mutate(crude_rate = round((Case_Count / Population) * 100000, digits = 2))

# generate plot visualization
AZ_catch_oral_cavity_white_plot <- oral_cavity_catch_white %>%
  filter(Year != "All") %>%
  ggplot(mapping = aes(y = Age_Adj_Rate, x = Year, color = Sex)) +
  geom_point() +
  scale_y_continuous(limits = c(0, 20)) +
  scale_x_discrete(labels = NULL) +
  labs(
    title = "Age-Adjusted Cancer Incidence Rates: Oral Cavity",
    subtitle = "for Catchment, years 1997-2017, White, Non-Hispanic",
    y = "Age Adjusted Rate per 100,000",
    x = "Year 1997-2017",
    caption = "Source: Arizona Cancer Registry, 1997-2017"
  ) +
  theme_solarized()

# prepare data table with variables Year, Male, Female, Total
AZ_catch_oral_cavity_white_table <- oral_cavity_catch_white %>%
  filter(Year != "All") %>%
  select(Year, Sex, Age_Adj_Rate) %>%
  spread(Sex, Age_Adj_Rate) %>%
  select(Year, Male, Female, All) %>%
  kable()

# catchment, white hispanic ----
# age adjusted incidence for cancer of the oral cavity
# read dataset to environment
# years 1997-2017
# cancer site: oral cavity
# white, Hispanic
# Cochise, Pima, Pinal, Santa Cruz, and Yuma counties only
# both male and female sex
oral_cavity_catch_hispanic <- read_csv("../../../data/raw/az_cancer_registry_incidence_oral_cavity_catchment_hispanic_by_year_20200604.csv",
  na = c("", "NA", "**", "*"),
  skip = 1,
  col_names = c(
    "Year",
    "Sex",
    "Case_Count",
    "Population",
    "Age_Adj_Rate",
    "95CI_min",
    "95CI_max"
  ),
  col_types = cols(
    Year = col_factor(),
    Sex = col_factor(),
    Case_Count = col_number(),
    Population = col_number(),
    Age_Adj_Rate = col_number(),
    "95CI_min" = col_number(),
    "95CI_max" = col_number()
  ),
)

glimpse(oral_cavity_catch_hispanic)

# calculate crude rate and add calculated variable
oral_cavity_catch_hispanic <- oral_cavity_catch_hispanic %>%
  mutate(crude_rate = round((Case_Count / Population) * 100000, digits = 2))

# generate plot visualization
AZ_catch_oral_cavity_plot_hispanic <- oral_cavity_catch_hispanic %>%
  filter(Year != "All") %>%
  ggplot(mapping = aes(y = Age_Adj_Rate, x = Year, color = Sex)) +
  geom_point() +
  scale_y_continuous(limits = c(0, 20)) +
  scale_x_discrete(labels = NULL) +
  labs(
    title = "Age-Adjusted Cancer Incidence Rates: Oral Cavity",
    subtitle = "for Catchment, years 1997-2017, White, Hispanic",
    y = "Age Adjusted Rate per 100,000",
    x = "Year 1997-2017",
    caption = "Source: Arizona Cancer Registry, 1997-2017"
  ) +
  theme_solarized()

# prepare data table with variables Year, Male, Female, Total
AZ_catch_oral_cavity_table_hispanic <- oral_cavity_catch_hispanic %>%
  filter(Year != "All") %>%
  select(Year, Sex, Age_Adj_Rate) %>%
  spread(Sex, Age_Adj_Rate) %>%
  select(Year, Male, Female, All) %>%
  kable()

# arizona, by race ----
# age adjusted incidence for cancer of the oral cavity grouped by race
# read dataset to environment
# years 1997-2017
# cancer site: oral cavity
# grouped by race
# all AZ counties
oral_cavity_race <- read_csv("../../../data/raw/az_cancer_registry_incidence_oral_cavity_race_20200604.csv",
  na = c("", "NA", "**", "*"),
  skip = 1,
  col_names = c(
    "Race",
    "Sex",
    "Case_Count",
    "Population",
    "Age_Adj_Rate",
    "95CI_min",
    "95CI_max"
  ),
  col_types = cols(
    Race = col_factor(),
    Sex = col_factor(),
    Case_Count = col_number(),
    Population = col_number(),
    Age_Adj_Rate = col_number(),
    "95CI_min" = col_number(),
    "95CI_max" = col_number()
  ),
)

glimpse(oral_cavity_race)

# calculate crude rate and add calculated variable
oral_cavity_race <- oral_cavity_race %>%
  mutate(crude_rate = round((Case_Count / Population) * 100000, digits = 2))

# generate plot visualization
AZ_oral_cavity_race_plot <- oral_cavity_race %>%
  filter(Race != "All") %>%
  ggplot(mapping = aes(x = Age_Adj_Rate, y = Race, fill = Sex)) +
  geom_bar(position = "dodge", stat = "identity", alpha = 0.75, color = "gray") +
  scale_x_continuous(limits = c(0, 20)) +
  labs(
    title = "Age-Adjusted Cancer Incidence Rates: Oral Cavity",
    subtitle = "Arizona total for years 1997-2017",
    x = "Age Adjusted Rate per 100,000",
    caption = "Source: Arizona Cancer Registry, 1997-2017"
  ) +
  theme_solarized()

# prepare data table
AZ_oral_cavity_race_table <- oral_cavity_race %>%
  select(Race, Sex, Age_Adj_Rate) %>%
  spread(Sex, Age_Adj_Rate) %>%
  select(Race, Male, Female, All) %>%
  kable()

# catchment, by race ----
# age adjusted incidence for cancer of the oral cavity grouped by race
# read dataset to environment
# years 1997-2017
# cancer site: oral cavity
# grouped by race
# Cochise, Pima, Pinal, Santa Cruz, and Yuma counties only
oral_cavity_race_catch <- read_csv("../../../data/raw/az_cancer_registry_incidence_oral_cavity_catchment_race_20200604.csv",
  na = c("", "NA", "**", "*"),
  skip = 1,
  col_names = c(
    "Race",
    "Sex",
    "Case_Count",
    "Population",
    "Age_Adj_Rate",
    "95CI_min",
    "95CI_max"
  ),
  col_types = cols(
    Race = col_factor(),
    Sex = col_factor(),
    Case_Count = col_number(),
    Population = col_number(),
    Age_Adj_Rate = col_number(),
    "95CI_min" = col_number(),
    "95CI_max" = col_number()
  ),
)

glimpse(oral_cavity_race_catch)

# calculate crude rate and add calculated variable
oral_cavity_race_catch <- oral_cavity_race_catch %>%
  mutate(crude_rate = round((Case_Count / Population) * 100000, digits = 2))

# generate plot visualization
AZ_catch_oral_cavity_race_plot <- oral_cavity_race_catch %>%
  filter(Race != "All") %>%
  ggplot(mapping = aes(x = Age_Adj_Rate, y = Race, fill = Sex)) +
  geom_bar(position = "dodge", stat = "identity", alpha = 0.75, color = "gray") +
  scale_x_continuous(limits = c(0, 20)) +
  labs(
    title = "Age-Adjusted Cancer Incidence Rates: Oral Cavity",
    subtitle = "Catchment total for years 1997-2017",
    x = "Age Adjusted Rate per 100,000",
    caption = "Source: Arizona Cancer Registry, 1997-2017"
  ) +
  theme_solarized()

# prepare data table
AZ_catch_oral_cavity_race_table <- oral_cavity_race_catch %>%
  select(Race, Sex, Age_Adj_Rate) %>%
  spread(Sex, Age_Adj_Rate) %>%
  select(Race, Male, Female, All) %>%
  kable()
```




| Site Group | ICD-O-3 Site (diagnosed 2001 to present) | ICD-O-3 Histology (Type) |
| :-------   | :------                                  | :-----                   |
| Lip | C000-C009 | excluding 9590-9989 |
| Tongue | C019-C029 | excluding 9590-9989 |
|  Salivary Gland  |  C079-C089 | excluding 9590-9989 |
|  Floor of Mouth  |  C040-C049 | excluding 9590-9989 |
|  Gum and Other Mouth  |  C030-C039, C050-C059, C060-C069 | excluding 9590-9989 |
|  Nasopharynx  |  C110-C119 | excluding 9590-9989 |
|   Tonsil  |  C090-C099 | excluding 9590-9989 |
| Oropharynx  |  C100-C109 | excluding 9590-9989 |
|  Hypopharynx  |  C129, C130-C139 | excluding 9590-9989 |
|  Other Oral Cavity and Pharynx  | C140, C142-C148 | excluding 9590-9989 |



## By Year and Sex


### Arizona 



#### All Races


The following chart and table show the age adjusted incidence rate for cancer of the oral cavity for:


- the entire state of Arizona
- years 1997-2017
- all races
- both female and male sex


```{r az_plot}
AZ_oral_cavity_incidence_plot
```


```{r az_table}
AZ_oral_cavity_incidence_table
```


#### White Non-Hispanic


The following chart and table show the age adjusted incidence rate for cancer of the oral cavity for:


- the entire state of Arizona
- years 1997-2017
- only the race listed as "White, Non Hispanic"
- both female and male sex


```{r white_plot}
AZ_oral_cavity_incidence_white_plot
```


```{r white_table}
AZ_oral_cavity_incidence_white_table
```

#### White Hispanic 


The following chart and table show the age adjusted incidence rate for cancer of the oral cavity for:


- the entire state of Arizona
- years 1997-2017
- only the race listed as "White, Hispanic"
- both female and male sex



```{r hisp_plot}
AZ_oral_cavity_incidence_hispanic_plot
```


```{r hisp_table}
AZ_oral_cavity_incidence_hispanic_table
```


### Five County Catchment 


#### Year and Sex


The following chart and table show the age adjusted incidence rate for cancer of the oral cavity for:


- only the five counties in the catchment; Cochise, Pima, Pinal, Santa Cruz, and Yuma
- years 1997-2017
- all races 
- all sexes 



```{r catch_plot}
AZ_catch_oral_cavity_plot
```


```{r catch_table}
AZ_catch_oral_cavity_table
```


#### White Non Hispanic 


The following chart and table show the age adjusted incidence rate for cancer of the oral cavity for:


- only the five counties in the catchment; Cochise, Pima, Pinal, Santa Cruz, and Yuma
- years 1997-2017
- only the race listed as "White, Non Hispanic" 
- all sexes 


```{r catch_white_plot}
AZ_catch_oral_cavity_white_plot
```


```{r catch_white_table}
AZ_catch_oral_cavity_white_table
```

#### White, Hispanic 


The following chart and table show the age adjusted incidence rate for cancer of the oral cavity for:


- only the five counties in the catchment; Cochise, Pima, Pinal, Santa Cruz, and Yuma
- years 1997-2017
- only the race listed as "White, Hispanic" 
- all sexes 


```{r catch_hisp_plot}
AZ_catch_oral_cavity_plot_hispanic
```


```{r catch_hisp_table}
AZ_catch_oral_cavity_table_hispanic
```


## By Race


### Arizona


The following chart and table show the age adjusted incidence rate for cancer of the oral cavity for:


- the entire state of Arizona
- all years 1997-2017
- all races 
- all sexes 


```{r race_plot}
AZ_oral_cavity_race_plot
```


```{r race_table}
AZ_oral_cavity_race_table
```

### Catchment


The following chart and table show the age adjusted incidence rate for cancer of the oral cavity for:


- only the five counties in the catchment; Cochise, Pima, Pinal, Santa Cruz, and Yuma
- all years 1997-2017
- all races 
- all sexes 


```{r race_catch_plot}
AZ_catch_oral_cavity_race_plot
```


```{r race_catch_table}
AZ_catch_oral_cavity_race_table
```
