---
title: "Pima County, AZ: Dashboard Describing the Cancer Burden"
output: 
  flexdashboard::flex_dashboard:
    storyboard: true
    social: menu
    theme: journal
---


```{r setup, include=FALSE}
# load packages
library(flexdashboard)
library(here)
library(tidyverse)
library(knitr)
library(ggthemes)
library(plotly)

# read data
pima_incidence_by_cancer <- read_rds("data/tidy/azdhs_pima_2013-2017_incidence_by_cancer.rds")
pima_incidence_by_race <- read_rds("data/tidy/azdhs_pima_2013-2017_incidence_by_race.rds")

# plot
plot_pima_incidence_by_cancer <- pima_incidence_by_cancer %>%
  group_by(Sex) %>%
  filter(Cancer != "All") %>%
  arrange(desc(Age_Adj_Rate)) %>%
  slice(1:5) %>%
  ggplot(mapping = aes(y = reorder(Cancer, Age_Adj_Rate), x = Age_Adj_Rate)) +
  geom_col(position = "dodge", alpha = 0.8) +
  facet_wrap("Sex") +
  theme_solarized() +
  theme(legend.position = "bottom") +
  labs(title = "Top 5 Incident Cancer for Pima County, AZ",
       subtitle = "2013-2017; All races; Grouped by sex",
       x = "Age Adjusted Rate per 100,000",
       y = "",
       caption = "Source: Arizona Cancer Registry Query Module")

# plot
plot_pima_incidence_by_race <-  pima_incidence_by_race %>%
  group_by(Race, Sex) %>%
  arrange(desc(Age_Adj_Rate)) %>%
  slice(1:5) %>%
  ggplot(mapping = aes(y = reorder(Sex, Age_Adj_Rate), x = Age_Adj_Rate, fill = Sex)) +
  geom_col(position = "dodge", alpha = 0.8) +
  facet_wrap("Race") +
  theme_solarized() +
  theme(legend.position = "bottom") +
  labs(title = "Top 5 Incident Cancer for Pima County, AZ",
       subtitle = "2013-2017; All races; Grouped by race and sex",
       x = "Age Adjusted Incidence Rate per 100,000",
       y = "",
       caption = "Source: Arizona Cancer Registry Query Module")
```


### Top 5 New Cancers for Pima County, AZ grouped by sex, for all races 


```{r incidence_by_cancer}
# show top five as plot 
ggplotly(plot_pima_incidence_by_cancer)
```


***


The chart shows the top five incident cancers for:


- male and female combined
- male only
- female only


Data shown is the five year average for years 2013-2017, for all races combined. 


Source: 


The Arizona Department of Health Services (ADHS) Indicator Based Information System for Public Health (IBIS-PH) system allows the public to query cancer rates, mortality rates and population estimates for Arizona. These public health data sets are intended to support evidenced-based decision making in Arizona to plan and improve service delivery, evaluate health care systems, and inform policy decisions. Other uses are not permissible. Available at http://healthdata.az.gov/query/module_selection/azcr/AzCRSelection.html


### New Cancers for Pima County, AZ grouped by race and by sex


```{r incidence_by_race}
ggplotly(plot_pima_incidence_by_race)
```


*** 


The chart shows the incident cancer for each race in Pima County, AZ grouped by sex. 


Data source: 


The Arizona Department of Health Services (ADHS) Indicator Based Information System for Public Health (IBIS-PH) system allows the public to query cancer rates, mortality rates and population estimates for Arizona. These public health data sets are intended to support evidenced-based decision making in Arizona to plan and improve service delivery, evaluate health care systems, and inform policy decisions. Other uses are not permissible. Available at http://healthdata.az.gov/query/module_selection/azcr/AzCRSelection.html


### Age adjusted incidence rates by age for Pima County, AZ


*** 


This panel shows a graph of the age adjusted incidence rates by age for Pima County, AZ


### Age adjusted incidence rates by year for Pima County, AZ


*** 


This panel shows a graph of the age adjusted incidence rates by age for Pima County, AZ


### Top 5 Cancer Deaths for Pima County, AZ grouped by sex


*** 


This panel shows a graph of the top 5 cancer deaths for Pima County, AZ


### Cancer Deaths for Pima County, AZ grouped by race and sex


*** 


This panel shows a graph of the cancer deaths for Pima County, AZ grouped by race and sex.


### 