---
title: "Pima County, AZ: Dashboard Describing the Cancer Burden"
output: 
  flexdashboard::flex_dashboard:
    storyboard: true
    social: menu
    theme: journal
    source: embed
---


```{r setup, include=FALSE}
# load packages
library(flexdashboard)
library(here)
library(tidyverse)
library(knitr)
library(ggthemes)
library(plotly)

# incidence by cancer and sex
# read data
pima_incidence_by_cancer <- read_rds("data/tidy/azdhs_pima_2013-2017_incidence_by_cancer.rds")

# plot
plot_pima_incidence_by_cancer <- pima_incidence_by_cancer %>%
  group_by(Sex) %>%
  filter(Cancer != "All") %>%
  arrange(desc(Age_Adj_Rate)) %>%
  slice(1:5) %>%
  ggplot(mapping = aes(y = reorder(Cancer, Age_Adj_Rate), x = Age_Adj_Rate)) +
  geom_col(position = "dodge", alpha = 0.8) +
  facet_wrap("Sex") +
  theme_solarized() +
  theme(legend.position = "bottom") +
  labs(title = "Top 5 Incident Cancer for Pima County, AZ",
       subtitle = "2013-2017; All races; Grouped by sex",
       x = "Age Adjusted Rate per 100,000",
       y = "",
       caption = "Source: Arizona Cancer Registry Query Module")

# incidence by race and sex
# read data
pima_incidence_by_race <- read_rds("data/tidy/azdhs_pima_2013-2017_incidence_by_race.rds")

# plot
plot_pima_incidence_by_race <-  pima_incidence_by_race %>%
  group_by(Race, Sex) %>%
  arrange(desc(Age_Adj_Rate)) %>%
  slice(1:5) %>%
  ggplot(mapping = aes(y = reorder(Sex, Age_Adj_Rate), x = Age_Adj_Rate, fill = Sex)) +
  geom_col(position = "dodge", alpha = 0.8) +
  facet_wrap("Race") +
  theme_solarized() +
  theme(legend.position = "bottom") +
  labs(title = "Top 5 Incident Cancer for Pima County, AZ",
       subtitle = "2013-2017; All races; Grouped by race and sex",
       x = "Age Adjusted Incidence Rate per 100,000",
       y = "",
       caption = "Source: Arizona Cancer Registry Query Module")

# incidence by age and sex
# read data
pima_incidence_by_age <- read_rds("data/tidy/azdhs_pima_2013-2017_incidence_by_age.rds")

# show top five as plot 
plot_pima_incidence_by_age <- pima_incidence_by_age %>%
  filter(Age_Group != "All") %>%
  group_by(Sex) %>%
  arrange(desc(Case_Count)) %>%
  slice(1:5) %>%
  ggplot(mapping = aes(y = reorder(Age_Group, Case_Count), x = Case_Count)) +
  geom_col(position = "dodge", alpha = 0.8) +
  facet_wrap("Sex") +
  theme_solarized() +
  theme(legend.position = "bottom") +
  labs(title = "Age Groups with Highest Number of New Cancer Cases",
       subtitle = "2013-2017; All races; Grouped by sex",
       x = "New Cases",
       y = "",
       caption = "Source: Arizona Cancer Registry Query Module")

# incidence by sex over time
# read data
# pima incidence over time by year ----
# read data
pima_incidence_by_year <- read_rds("data/tidy/azdhs_pima_1995-2017_incidence_by_year.rds")

# show plot 
plot_pima_incidence_by_year <- pima_incidence_by_year %>%
  ggplot(mapping = aes(y = Age_Adj_Rate, x = Year, group = Sex, color = Sex)) +
  geom_line() +
  theme_solarized() +
  ylim(c(200, 600)) +
  theme(legend.position = "bottom") +
  labs(title = "Age Adjusted Incidence Rate of New Cancers in Pima County, AZ",
       subtitle = "1995-2017; All races combined; Grouped by sex",
       x = "Year",
       y = "Age Adjusted Incidence Rate per 100,000",
       caption = "Source: Arizona Cancer Registry Query Module")


# mortality by cancer and sex
# read data
pima_mortality_by_cancer <- read_rds("data/tidy/seer_stat_pima_mortality_2013-2017.rds")

# determine top five cancers for each sex grouping 
# filter out "system"
# and save to value for use later
cancer_to_exclude <- str_subset(pima_mortality_by_cancer$cancer, " System")

# show plot
plot_pima_mortality_by_cancer <- pima_mortality_by_cancer %>% 
  group_by(sex) %>%
  filter(cancer != "Miscellaneous Malignant Cancer" &
           cancer != "Colon excluding Rectum" &
           cancer != "All Causes of Death" & 
           cancer != "All Malignant Cancers" & 
           !(cancer %in% cancer_to_exclude)) %>%
  select(sex, cancer, age_adjusted_rate) %>%
  arrange(desc(age_adjusted_rate)) %>%
  slice(1:5) %>%
  ggplot(mapping = aes(x = age_adjusted_rate, y = reorder(cancer, age_adjusted_rate))) +
  geom_col(position = "dodge", alpha = 0.8) +
  labs(title = "Top 5 Cancer Deaths in Pima County, AZ",
       subtitle = "Five year average 2013-2017; All races",
       x = "Age Adjusted Mortality Rate per 100,000",
       y = "",
       caption = "Source: Surveillance, Epidemiology, and End Results (SEER) Program (www.seer.cancer.gov)") +
  facet_wrap("sex") +
  theme_solarized()

# mortality by race and sex
# read data
pima_mortality_by_race <- read_rds("data/tidy/seer_stat_pima_mortality_2013-2017_by_race.rds")

# show plot
plot_pima_mortality_by_race <- pima_mortality_by_race %>% 
  group_by(race, sex) %>%
  ggplot(mapping = aes(x = age_adjusted_rate, y = reorder(sex, age_adjusted_rate), fill = sex)) +
  geom_col(position = "dodge", alpha = 0.8) +
  labs(title = "Top 5 Cancer Deaths in Pima County, AZ",
       subtitle = "Five year average 2013-2017; All races",
       x = "Age Adjusted Mortality Rate per 100,000",
       y = "",
       caption = "Source: Surveillance, Epidemiology, and End Results (SEER) Program (www.seer.cancer.gov)") +
  facet_wrap("race") +
  theme_solarized() +
  theme(legend.position = "bottom")

# mortality by age and sex
# read data
pima_mortality_by_age <- read_rds("data/tidy/seer_stat_pima_mortality_2013-2017_by_age.rds")

# show plot
plot_pima_mortality_by_age <- pima_mortality_by_age %>%
  group_by(sex) %>%
  arrange(desc(age_adjusted_rate)) %>%
  slice(1:5) %>%
  ggplot(mapping = aes(x = age_adjusted_rate, y = reorder(age_group, age_adjusted_rate))) +
  geom_col(position = "dodge") +
  facet_wrap("sex") +
  labs(title = "Age Groups with Most Cancer Deaths in Pima County, AZ",
       subtitle = "Five year average 2013-2017; All races combined",
       x = "Age Adjusted Mortality Rate per 100,000",
       y = "",
       caption = "Source: Surveillance, Epidemiology, and End Results (SEER) Program (www.seer.cancer.gov)") +
  theme_solarized() +
  theme(legend.position = "bottom")

# mortality by sex over time
# read data

# show plot
```


### Top 5 New Cancers for Pima County, AZ grouped by sex, for all races 


```{r incidence_by_cancer}
# show top five as plot 
ggplotly(plot_pima_incidence_by_cancer)
```


***


The chart shows the top five incident cancers for:


- male and female combined
- male only
- female only


Data shown is the five year average for years 2013-2017, for all races combined. 


Source: 


The Arizona Department of Health Services (ADHS) Indicator Based Information System for Public Health (IBIS-PH) system allows the public to query cancer rates, mortality rates and population estimates for Arizona. These public health data sets are intended to support evidenced-based decision making in Arizona to plan and improve service delivery, evaluate health care systems, and inform policy decisions. Other uses are not permissible. Available at [http://healthdata.az.gov/](http://healthdata.az.gov/query/module_selection/azcr/AzCRSelection.html)


### New Cancers for Pima County, AZ grouped by race and by sex


```{r incidence_by_race}
ggplotly(plot_pima_incidence_by_race)
```


*** 


The chart shows the five year average (2013-2017) of incident cancer for each race in Pima County, AZ grouped by sex. 


Data source: 


The Arizona Department of Health Services (ADHS) Indicator Based Information System for Public Health (IBIS-PH) system allows the public to query cancer rates, mortality rates and population estimates for Arizona. These public health data sets are intended to support evidenced-based decision making in Arizona to plan and improve service delivery, evaluate health care systems, and inform policy decisions. Other uses are not permissible. Available at [http://healthdata.az.gov/](http://healthdata.az.gov/query/module_selection/azcr/AzCRSelection.html)


### Age adjusted incidence rates by age for Pima County, AZ


```{r incidence_by_age}
ggplotly(plot_pima_incidence_by_age)
```


*** 


This panel shows a graph of the age groups with the highest number of new cases for the five year period from 2013 - 2017; for Pima County, AZ.


Data source: 


The Arizona Department of Health Services (ADHS) Indicator Based Information System for Public Health (IBIS-PH) system allows the public to query cancer rates, mortality rates and population estimates for Arizona. These public health data sets are intended to support evidenced-based decision making in Arizona to plan and improve service delivery, evaluate health care systems, and inform policy decisions. Other uses are not permissible. Available at [http://healthdata.az.gov/](http://healthdata.az.gov/query/module_selection/azcr/AzCRSelection.html)


### Age adjusted incidence rates by year for Pima County, AZ


```{r incidence_by_year}
ggplotly(plot_pima_incidence_by_year)
```


*** 


This panel shows a graph of the age adjusted incidence rates by for Pima County, AZ over time 


Data source: 


The Arizona Department of Health Services (ADHS) Indicator Based Information System for Public Health (IBIS-PH) system allows the public to query cancer rates, mortality rates and population estimates for Arizona. These public health data sets are intended to support evidenced-based decision making in Arizona to plan and improve service delivery, evaluate health care systems, and inform policy decisions. Other uses are not permissible. Available at [http://healthdata.az.gov/](http://healthdata.az.gov/query/module_selection/azcr/AzCRSelection.html)



### Top 5 Cancer Deaths for Pima County, AZ grouped by sex


```{r mortality_by_cancer}
ggplotly(plot_pima_mortality_by_cancer)
```


*** 


This panel shows a graph of the top 5 cancer deaths for Pima County, AZ


Data source: 


Surveillance, Epidemiology, and End Results (SEER) Program ([www.seer.cancer.gov](www.seer.cancer.gov)) SEER*Stat Database: Mortality - All COD, Aggregated With County, Total U.S. (1990-2017) <Katrina/Rita Population Adjustment> - Linked To County Attributes - Total U.S., 1969-2018 Counties, National Cancer Institute, DCCPS, Surveillance Research Program, released December 2019.  Underlying mortality data provided by NCHS ([www.cdc.gov/nchs](www.cdc.gov/nchs)).


### Cancer Deaths for Pima County, AZ grouped by race and sex


```{r mortality_by_race}
ggplotly(plot_pima_mortality_by_race)
```


*** 


This panel shows a graph of the cancer deaths for Pima County, AZ grouped by race and sex.


Data source: 


Surveillance, Epidemiology, and End Results (SEER) Program (www.seer.cancer.gov) SEER*Stat Database: Mortality - All COD, Aggregated With County, Total U.S. (1990-2017) <Katrina/Rita Population Adjustment> - Linked To County Attributes - Total U.S., 1969-2018 Counties, National Cancer Institute, DCCPS, Surveillance Research Program, released December 2019.  Underlying mortality data provided by NCHS (www.cdc.gov/nchs).


### Age Groups with highest number of Cancer Deaths for Pima County, AZ 


```{r mortality_by_age}
ggplotly(plot_pima_mortality_by_age)
```


***


The age groups with the highest number of cancer deaths in Pima County, AZ; presented as an age adjusted rate 


Data source: 


Surveillance, Epidemiology, and End Results (SEER) Program (www.seer.cancer.gov) SEER*Stat Database: Mortality - All COD, Aggregated With County, Total U.S. (1990-2017) <Katrina/Rita Population Adjustment> - Linked To County Attributes - Total U.S., 1969-2018 Counties, National Cancer Institute, DCCPS, Surveillance Research Program, released December 2019.  Underlying mortality data provided by NCHS (www.cdc.gov/nchs).


### Cancer Deaths Over Time in Pima County, AZ


... in progress ...


***


Describe the visualization 


### Population Demographics: Total population in Pima County 


... in progress ...


***


Describe the visualization


### Population Demographics: Sex and Age 


... in progress ...


***


Describe the visualization


### Population Demographics: Sex and Race 


... in progress ...


***


Describe the visualization


### Population Demographics: Education 


... in progress ...


***


Describe the visualization


### Population Demographics: Income 


... in progress ...


***


Describe the visualization


### Screening:


... in progress ...


***


Describe the visualization


### Health Behavior


... in progress ...


***


Describe the visualization


